From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2017-11-01 paper/paper.Rmd
here we go

csl: ecology-letters.csl
bibliography: refs.bib
output: rticles::elsevier_article  
layout:  5p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: "\\newcommand{\\ud}{\\mathrm{d}}"

@@

What is noise? Where does it come from?  It is important to bear
in mind that whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
Simply put, *noise is all that we leave out of our model*.  Early
quantitative ecological models first formulated at the population
level as differential or difference equations: deterministic models
that seek to capture the essential features of a system while
neglecting any random component as mere noise [@Coulson2004].
Advances in both theoretical understanding and the advent of widely
available and efficient numerical methods have helped drive research
towards increasingly more explicit descriptions of how stochastiticity
enters into quantitative models [@Black2012; @Ovaskainen2010].

@@

It is possible to simulate such a process exactly using Gillespie's
exact stochastic simulation algorithm [SSA; @Gillespie1977], which
simulates each individual birth or death as a separate event. Figure 1
uses this algorithm to compare simulations of the Levins' patch model;
see Appendix A for example code.  In the Levins'  model,
$n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.  Figure
1 shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$.

@@

where $x = \langle n \rangle$, the average or expected population
size of an ensemble of replicates, assuming the transitions (+/- 1
in this case) are much smaller than the system size ($N$ in the
Levins model).  Note that the dynamics for the average population
match the population-level equation^[up to fluctuations of order
$\tfrac{1}{N}$. We can actually both observe and successfully
approximate this higher-order correction from additional terms in
the system size expansion to recover the phenomenon of stochastic
inflation [@Gonzalez2002], where the average population size can
be larger or smaller than predicted by the deterministic model --
see Appendix A for examples in simulations compared to predictions
of the system size expansion].

Note that like the equation for the mean, we have a dynamic equation
for the time-evolution of the variance in Eq \eqref{bd}.  We can
solve these equations at steady-state by setting both equations
zero. For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$,
we find the average number of occupied patches 
$\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of 
$\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally
to the system size, $N$, which explains the standard deviation
increases only as $\sqrt{N}$ while the steady state itself increases
in direct proportion to $N$, resulting in the smaller relative size
of fluctuations in the larger system in Figure 1.  We can also see
that different ways of partitioning the same macroscopic equation
into components of birth and death will create the same mean dynamics
but with different levels of noise.  For instance, in a logistic
model with $b = rn$ and $d = rn^2/N$, we instead find the variance
at steady state is $\sigma^2 = K$; significantly larger than in the
Levins model.

@@

The importance of $\tau_c$, or the degree of auto-correlation, which
emerges from our derivation above, has also been and continues to be a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work rarely often assumed environmental noise would be
uncorrelated as a mathematical convenience.  As the derivation of
Eq \eqref{env_noise} makes clear, for a continuous-time process,
the relevant metric is the timescale of auto-correlation in the
environmental noise process relative to the timescale (auto-correlation)
of the population dynamics.  The Gaussian noise that emerges out
of the system-size approximation is always auto-correlated to some
degree -- if we measure our environment at consecutive times
infinitesimally far apart, we expect nearly identical values.  At
intervals much longer than the correlation time, consecutive
measurements are effectively independent (white noise). 
The degree of auto-correlation in  noise is all about _relative_ timescales.

@@

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals
within the population, or at least within a particular age or stage
class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically
as a distribution rather than a fixed value, making it function as
another form of random variation in a model [@Coulson2001;
@Schreiber2010].  *Spatial heterogeneity* can likewise be reflected
in stochastic parameters of a spatially implicit model, or as
spatially varying environmental stochasticity in a spatially explicit
model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
interaction between individual or spatial variation and other sources
of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
@Schreiber2017].  *Observational error*: Observational error acts
much like another form of stochasticity, arising from sampling
effects or limitations of instruments, etc. Yet unlike the sources
described above, this uncertainty does not impact the actual state
of our system, and thus does not give rise to any additional phenomena
or information about that process, though it can obscure it
[@Knape2011; @deValpine2002].

@@

A natural focus of much research into stochastic ecological systems
has been on understanding the implications for persistence of
individual and coexistence of interacting species. Demographic noise
in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise.  However, in most
models it is possible to define a notion of quasi-stationary 
distribution concentrated around the deterministic attractor,
and to prove the probability of extinction decreases exponentially with 
system size (e.g. see @Ovaskainen2010; @Schreiber2017 also reviews 
more recent literature which has established these results more generally). 
Generalizations about persistence and stable coexistence of interacting
species in stochastic models are difficult due to complex interactions
between demographic and environmental noise, population structure,
spatial heterogeneity, magnitude of noise, degree of autocorrelation and so forth
[e.g. @Higgins1997; @Coulson2001; @Melbourne2008; @Hart2016];
indeed care must be taken even in defining the approriate notion of 
persistence [e.g. @Schreiber2006].  Instead of these long-term limits,
here we will focus on a few examples of stochastic phenomena 
that can emerge on shorter time scales and provide some intuition for 
how the interaction between stochastic an nonlinear dynamics can 
create these surprising results.

@@

Noise can provide novel information in other ways as well.  Stochastic
fluctuations act like a continual miniature perturbation experiment,
constantly pushing a system away from an equilibrium to reveal its
behavior elsewhere in state space. Observing the size and
auto-correlation of these fluctuations thus provides insight on the
stability of an equilibrium point, (as can be seen, for instance
in the fluctuation-dissipation result we found for generic birth
death models).  This observation has been exploited with particular
success as a potential *early warning sign* that a system is loosing
stability; possibly due to an approaching catastrophic transition
or saddle-node bifurcation.  This pattern, known as critical slowing
down (in reference to the auto-correlation time in the physics
literature) was first proposed an an early warning sign for ecological
systems by @Wissel1984, but has received considerable empirical as
well as theoretical attention only more recently [@Scheffer2009;
@Scheffer2015].  Increases in two of these early warning indicators,
variance and auto-correlation, are illustrated along with the
timeseries of a simulation of the dynamics of yeast population
growth in Figure 4.  This simulation is based on a mechanistic model
developed by @Dai2012 that closely matches the population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cervisiae_ on a sucrose medium. Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation, which is accompanied
by the patterns of increasing variance and increasing auto-correlation
seen empirically in lab populations [@Dai2012] as well as the
numerical simulation shown in Figure 4.

@@

<!-- FIXME Fragment, also tie back to flickering results, move some of that here-->

Challenges remain in distinguishing these patterns from other causes [@Boettiger2013a] which will ultimately require more domain-specific information  [@Boettiger2013b].

>>>>

csl: ecology-letters.csl
bibliography: refs.bib
output: rticles::elsevier_article  
layout:  review # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \linenumbers

@@

What is noise? Where does it come from?  It is important to bear
in mind that whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
Simply put, *noise is all that we leave out of our model*.  Early
quantitative ecological models first formulated at the population
level as differential or difference equations: deterministic models
that seek to capture the essential features of a system while
neglecting any random component as mere noise [@Coulson2004].
Advances in both theoretical understanding and the advent of widely
available and efficient numerical methods have helped drive research
towards increasingly more explicit descriptions of how stochasticity
enters into quantitative models [@Black2012; @Ovaskainen2010].

@@

It is possible to simulate such a process exactly using Gillespie's
exact stochastic simulation algorithm [SSA; @Gillespie1977], which
simulates each individual birth or death as a separate event. Figure 1
uses this algorithm to compare simulations of the Levins' patch model;
see Appendix A for example code.  In the Levins'  model,
$n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.

Figure 1 shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$.

@@

where $x = \langle n \rangle$, the average or expected population
size of an ensemble of replicates, assuming the transitions (+/- 1
in this case) are much smaller than the system size ($N$ in the
Levins model).  The dynamics for the average population
match the population-level equation.^[up to fluctuations of order
$\tfrac{1}{N}$. We can actually both observe and successfully
approximate this higher-order correction from additional terms in
the system size expansion to recover the phenomenon of stochastic
inflation [@Gonzalez2002], where the average population size can
be larger or smaller than predicted by the deterministic model --
see Appendix A for examples in simulations compared to predictions
of the system size expansion].

The change in variance can be seen
as a balance between the stochastic contribution, $b(x) + d(x)$
and the deterministic contribution, $\partial_x \left(b(x) - d(x)\right)$. 
The deterministic contribution is the slope of the macroscopic
equation, that is, the eigenvalue indicating the stability of the
equilibrium point.  For stable points, this value must be negative
and precisely balance the strictly positive stochastic contribution 
of $b(x) + d(x)$ (the greater the rate of stochastic events involved, 
the greater the fluctuations).  This balance at steady-state is known as the
fluctuation-dissipation theorem [@vanKampen2007], an will prove a useful
expression in understanding stochastic dynamics.  Because Eq \eqref{bd}
also describes the time-evolution of these fluctuations, we can similarly
solve for the correlation between two points separated by time $\tau$.

For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$,
we find the average number of occupied patches 
$\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of 
$\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally
to the system size, $N$, which explains the standard deviation
increases only as $\sqrt{N}$ while the steady state itself increases
in direct proportion to $N$, resulting in the smaller relative size
of fluctuations in the larger system in Figure 1.  We can also see
that different ways of partitioning the same macroscopic equation
into components of birth and death will create the same mean dynamics
but with different levels of noise.  For instance, in a logistic
model with $b = rn$ and $d = rn^2/N$, we instead find the variance
at steady state is $\sigma^2 = K$; significantly larger than in the
Levins model.

@@

The importance of $\tau_c$, or the degree of auto-correlation, which
emerges from our derivation above, has also been and continues to be a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work rarely often assumed environmental noise would be
uncorrelated as a mathematical convenience.  As the derivation of
Eq \eqref{env_noise} makes clear, for a continuous-time process,
the relevant metric is the timescale of auto-correlation in the
environmental noise process relative to the auto-correlation timescale
of the population dynamics.  The Gaussian noise that emerges out
of the system-size approximation is always auto-correlated to some
degree -- if we measure our environment at consecutive times
infinitesimally far apart, we expect nearly identical values.  At
intervals much longer than the correlation time, consecutive
measurements are effectively independent (i.e. white noise). 
The degree of auto-correlation in  noise is all about _relative_ timescales.

@@

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals
within the population, or at least within a particular age or stage
class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically
as a distribution rather than a fixed value, making it function as
another form of random variation in a model [@Coulson2001;
@Schreiber2010].  *Spatial heterogeneity* can likewise be reflected
in stochastic parameters of a spatially implicit model, or as
spatially varying environmental stochasticity in a spatially explicit
model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
interaction between individual or spatial variation and other sources
of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
@Schreiber2017].  *Observational error* acts
much like another form of stochasticity, arising from sampling
effects or limitations of instruments, etc. Yet unlike the sources
described above, this uncertainty does not impact the actual state
of our system, and thus does not give rise to any additional phenomena
or information about that process, though it can obscure it
[@Knape2011; @deValpine2002].

@@

A natural focus of much research into stochastic ecological systems
has been on understanding the implications for persistence of
individual and coexistence of interacting species. Demographic noise
in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise.  However, in most
models it is possible to define a notion of quasi-stationary 
distribution concentrated around the deterministic attractor,
and to prove the probability of extinction decreases exponentially with 
system size (e.g. see @Ovaskainen2010; @Schreiber2017 also reviews 
more recent literature which has established these results more generally). 
Generalizations about persistence and stable coexistence of interacting
species in stochastic models are difficult due to complex interactions
between demographic and environmental noise, population structure,
spatial heterogeneity, magnitude of noise, degree of auto-correlation and so forth
[e.g. @Higgins1997; @Coulson2001; @Melbourne2008; @Hart2016];
indeed care must be taken even in defining the appropriate notion of 
persistence [e.g. @Schreiber2006].  Instead of these long-term limits,
here we will focus on a few examples of stochastic phenomena 
that can emerge on shorter time scales and provide some intuition for 
how the interaction between stochastic an nonlinear dynamics can 
create these surprising results.

@@

Noise can provide novel information in other ways as well.  Stochastic
fluctuations act like a continual miniature perturbation experiment,
constantly pushing a system away from an equilibrium to reveal its
behavior elsewhere in state space. Observing the size and
auto-correlation of these fluctuations thus provides insight on the
stability of an equilibrium point, (as can be seen, for instance
in the fluctuation-dissipation result we found for generic birth
death models).  This observation has been exploited with particular
success as a potential *early warning sign* that a system is loosing
stability; possibly due to an approaching catastrophic transition
or saddle-node bifurcation.  This pattern, known as critical slowing
down (in reference to the auto-correlation time in the physics
literature) was first proposed an an early warning sign for ecological
systems by @Wissel1984, but has received considerable empirical as
well as theoretical attention only more recently [@Scheffer2009;
@Scheffer2015].  Increases in two of these early warning indicators,
variance and auto-correlation, are illustrated along with the
timeseries of a simulation of the dynamics of yeast population
growth in Figure 4.  This simulation is based on a mechanistic model
developed by @Dai2012 that closely matches the population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cerevisiae_ on a sucrose medium. Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation, which is accompanied
by the patterns of increasing variance and increasing auto-correlation
seen empirically in lab populations [@Dai2012] as well as the
numerical simulation shown in Figure 4.

@@

The challenge of identifying warning signs that a system is approaching
a tipping point has proven a rich area for applications of stochastic
processes.  We have already seen how in addition to the patterns of
increasing variance and auto-correlation, our survey of stochastic phenomena
has highlighted the increased potential to observe stochastic switching,
or flickering, as the system approaches the tipping point.  Challenges remain
in distinguishing these patterns from other causes [@Boettiger2013a].
For instance, as we have seen, variance in a system can increase due to
external drivers without changing the auto-correlation, and both variance
and auto-correlation can increase if a system loses stability without
approaching a saddle-node bifurcation, but will not show flickering.
The greater our suite of stochastic indicators, the better our ability
to pin down the processes involved.  Such examples suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, in which a stable node becomes
a limit cycle.

Most importantly, the framing of stochastic processes as indicators of 
changes in an ecosystem that might otherwise go unnoticed has proven a
rich catalyst for bringing these concepts beyond the realm of abstract
theory. A rich array of experimental and observational tests of early
warning signals [_i.e._ @Carpenter2011; @Dai2012; and many others, _e.g._ see @Scheffer2015 review] has served no merely to illustrate these phenomena can be seen
in laboratory and field ecosystems, but also to spark suggestions of
additional possible indicators and surface a wide array of 
practical [@Boettiger2013b; @Scheffer2015], statistical [e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015; @Wysham2010] challenges to such application.  This
creates a two-way dialog between empirical and theoretical work that does
much to advance the field as a whole.
